Monitoring and Projecting Global Hunger: Are We On Track?

### 2021-07-16 docs/response_to_reviewers.md

---
created: 2021-06-10T10:04:08-0400
modified: 2021-07-16T16:31:14-0400
title: Response to Reviewers
header-includes:
 - \hypersetup{colorlinks=true}
 - \hypersetup{urlcolor=blue}
---

**We thank the reviewer for the detailed and thoughtful feedback, which forced us to engaged more with issues of spatial scale, household vs population food security, as well as a more broad body of literature thn we originally examined.  We believe that the paper is now much improved and will be of interest to a wider body of researchers, with more clear explanations of the methodlogy and spatial scale, and more detailed discussions of issues like population, urbanization, and climate.**

@@

**We thank you for this detailed feedback, and we agree that the approach to modeling the impacts of climate on food security is too simple to capture all of the pathways by which the climate affects food security.  However, we are in many ways constrained by our survey data, which is only available at the annual temporal scale and only reported at the national spatial scale.  This makes it much harder to trace the impacts of climate on food security and necessitates the smoothing of our climate data.  The papers you cite by Niles, Blekking, and Shukla all seem to have indicators of food security available at fine spatial scales (Niles uses DHS clusters/community GPS points, Blekking uses household GPS points) and fine temporal scales (Niles uses monthly data, Shukla uses Dekads).**

**We also note that it is important to consider the scope of our analysis.  Weather shocks as well as labor and food prices certainly explain a lot of the variation in food security between households and across seasons.  However, the scope of this analysis is to explore drivers of food insecurity across the globe and over decadal timeframes.  We are especially interested in modeling future trends in food security at these broad spatial and temporal scales.  We therefore use GCM projections, which are useful insofar as they are indicative of shifting baselines, but are unable to anticipate actual droughts or floods that would impact food insecurity in 2030.**

**Given these considerations of how the scale of our data (annual, national sampling) and scope of our analysis (modeling food insecurity globally and out to 2030) makes it infeasible to explore the effect of climate on food insecurity beyond shifting baselines, we do not think that using other datasets like ERA-5 or MERRA-2 would add much value. (Also, the CHIRPS dataset, while excellent, is not truly global and unavailable for certain latitudes, so couldnt be applied to this global analysis).  However, you raise very important considerations in your feedback and we therefore make two improvements to the paper: (1) in addition to the survey year's rainfall totals and precipitation averages, we add a feature for the previous year's total rainfall and mean temperature, and additionally (2) we add a paragraph to the discussion highlighting many of the points you raise and further outlining the considerations that led to the analytical approach we took.**

@@

Third, Random Forrest models suffer from training data bias. If the underlying training data if flawed, the Random Forest Model has no way of identifying this. What quality control is used to determine if your survey data is correct? One way to address this is to compare the Gallup data with other sources like DHS or LSMS surveys from the same year. Do they both measure similar national-level food security outcomes? I am also curious to know how your models preform with simple linear or logistic regression? Do you get the same results? I recognize Random Forest are nonparametric, but I would be keen to see how linear models preform with controls. Also, what happens when you trade out wasting/stunting as your predicted outcome versus the FIES-based metric?

**Unforuntately, the Gallup World Poll (GWP) is currently the only source of nationally-respresntative data on the FIES, as this metric is not collected by surveys like the DHS and LSMS (although as the FIES has been adopted as an SDG indicator, this is changing, see [here](https://userforum.dhsprogram.com/index.php?t=msg&th=7832&start=0&)).  Thus, we cannot use those datasets to validate our Random Forest model.  However, we note that we do take care to conduct cross validtion at the country level.  We leave out all surveys a given country and then evaluate the model's performance on that country, and find our model it still performs favorably.**

**Additionally, we have updated the paper to now model the FIES using LASSO regression (in the supplement), and note that, compared to linear regression, the random forest model performs quite well.  We also now add some discussion around the importance of accounting for interactions between covariates based on a comparison between random forest and linear regression.**

@@

Fourth, the spatial nature of your modeling is confusing and not clear. Are metrics averaged at the sub-national scale (if so what level of analysis and for which country)? How is degree of urbanization quantified (e.g. are Worldpop pixels masked at some threshold?)? Is the Gallup representative at the national level? District-level? What controls are used to ensure that the Gallup data is a representative sample (Not even explained in the Supplemental Materials)? Your methodology (including how you address issues of spatial scale) needs to be more clearly described in the Main Text and the Supplemental Material before this paper can be considered for publication.

**To add more clarity, we have changed the phrasing in most places from "subnational" to "administrative area", as this might improve some of the confusion around the spatial scale of our analysis.  We have also added a paragraph to the methods section giving a high-level overview of the analysis, clearly stating the scale of the study (year x admin area).  The Gallup data is designed to be representative at the national level through a weighting schema based on demographic characterstics, and we used the same methodology to get estimates at the administrative area level.  Gallup records whether individual respondents are in urban or rural areas based on the individuals own reporting.  We then use global gridded data from Jones and O'Neill (2016) published in _Environmental Research Letters_ to estimate the percent of the population in urban areas for each administrative area through the year 2030.**

@@

Pg. 7, Ln 19 - 30: This results is in contrast to FEWS NET outlook that the those at risk to famine rose by 85 million people from 2015 - 2019 (I can't find the brief, but here is an article with the numbers: https://www.sciencemag.org/news/2020/04/how-team-scientists-studying-drought-helped-build-world-s-leading-famine-prediction). This was before cofid. Why do your numbers differ?

**This is an interesting point. We note that the paragraph you take issue with is specifically focusing on the period of 2010 to 2030, and our figures, which include the range 2015-2019, show increases in severe and moderate food insecurity over this period, jibing with CHG's findings.  We also took some effort to compare our results to other reports of trends in food security, focusing mostly on the metric of undernourishment in the FAO's State of Food Insecurity report.  We now have a reference to the science news article you mentioned.**

Pg. 9, 3.2: You have sub-national data, and you make assumptions that the characteristics of two similar households have the same food security outcomes, yet you present your findings at national level. An increasing body of research (e.g. ref) is moving away from national-level estimates as they mask the true nature of food security outcomes and are not relevant for policy maker or targeted interventions. Please justify these findings given the high degree of spatiotemporal heterogeneity in food security outcomes.

**First, we share some of your concerns about the inadequacy of national-level food security estimates, which is precisely why we chose to disaggregate our analysis to the admin area level.  We further note that our data was originally available at the national scale, and we used demographic data to disaggregated it (See section 2.1).  Disaggregating to the admin 1 area is possible with our data, but a household or individual-level analysis really is not.**

**We also disagree that "national-level estimates... are not relevant for policy makers".  While local practitioners focus on differences in food security between inidivudals or households, many mulitnational practitions pay close attention to country-level (or admin-level) estimates of food security.  We note the FAO's State of Food Insecurity (SOFI) reports, which are widely read, as well as the fact that the SDG targets, which are key metrics for policymakers and which include the FIES for SDG2, are monitored at the national level (see: https://sdg-tracker.org/zero-hunger).  We discuss both the SOFI estimates and the relevance to SDG2 throughtout the paper.  Moreover, as we mention in the acknowledgments, this project began out of a specific demand from an international policymaking organization.  So we feel that our analysis meets a clear stated need by policymakers, and is conducted at the appropriate spatial scale given the data that we have.**

### 2021-07-16 docs/old/modeling_fies.tex
resubmission

%\maketitle
\begin{abstract}
This paper presents the first global picture of food security at an administrative area, subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.  Using microdata from 75 countries and filling gaps using machine learning, we find significant heterogeneity in levels of food insecurity around the world.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of severe food insecurity are declining, resulting in global decreases in the total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, we predict it will continue to increase through the end of the 2020s.  Overall, we conclude that current trends in development and demographic change will still leave a large share of the world's population still experiencing hunger by 2030.  \end{abstract}

@@

This paper contributes to shifting the research frontier by filling these two gaps in terms of spatial coverage and forecasts by developing a machine learning model of FIES-based food insecurity rates based on covariates that are available for territories where actual survey data are not available. This model is then employed to obtain forecasts to 2030 and assess likely future trends in food insecurity. We take advantage of individual-level characteristics collected in FAO's FIES surveys to push the analysis beyond the country level to a sub-national level. To our knowledge, this is the first effort to model FIES-based food insecurity at that scale and make projections to 2030.

@@

\section{Methods}
\subsection{Disaggregation}
The data on the FIES collected by Gallup on behalf of FAO is collected annually within a country, and is therefore at the country-year level.  This data also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  In order to disaggregate the data to the administrative area level, we use a similar methodology and calculate post-stratification weights.  We use year-specific administrative area estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each administrative area across all individuals.  For subnational distributions of age and gender, we rely on gridded population data.  While there are many different gridded population datasets, they often take different approaches to disaggregating population data, are available at different temporal resolutions, and have varying availability of demographic data \cite{Ley2019Sep, friedman2001elements, Tuholske2019Aug}.  We therefore chose to use the WorldPop dataset, which has demographic data on age and sex and is available annually \citep{Tatem2017}.  For rates of urbanization by administrative area, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}, and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.  This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.

\subsection{Covariates}
We model rates of moderate and severe food insecurity as a function of several covariates related to human development, nutrition, health, infrastructure, income levels and distribution, and the environment at the administrative area level (See Table \ref{tab:covars}).  We employ covariates for which projections are available up to year 2030 and, in most cases, also have administrative area spatial resolution.  For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.

Many of the variables require the harmonization of subnational historical data with projections at the national level.  These include mean years of schooling, GDP per capita, as well as population. We use observed trajectories in the historical distribution of variables among administrative areas within a country to predict the future distribution of GDP, population, and schooling among these areas and disaggregate national-level future projections.

For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each administrative area to model these and to obtain predictions for variables up to 2030. We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.  This method is used to obtain forecasts of the rates of stunting and wasting up to the year 2030.

@@

\subsection{Modeling}
We fit two models for all administrative areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.  A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.  By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}. Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.  We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.

@@

When performing cross validation, it is typical to select observations at random.  However, in our training data we have multiple highly similar observations from each country across years and administrative areas, and we extend our model predictions to countries that were not found at all in our training data.  Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.  Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.  By this validation metric, we find that our models perform well, with an $R^2$ of 0.792 and mean average error (MAE) of 0.071 for moderate food insecurity and an $R^2$ of 0.746 and a MAE of 0.041 for severe food insecurity.  After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.945 and a MAE of 0.018 for severe food insecurity.

@@

\subsection{Food Insecurity Across Space}
We find substantial heterogeneity in the global distribution of severe food insecurity across countries and administrative areas (See Fig. \ref{fig:map}).  For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with 20\% of people over the corresponding threshold in at least one administrative area in every country except Gabon, Djibouti and Equatorial Guinea.  Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.  In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10\% and 15\% in northern Brazil, many central Asian and middle-eastern countries, as well as India and Indonesia.

@@

We also find substantial within-country variation in rates of moderate and severe food insecurity in many countries.  In Brazil, for example, northern regions are much more food insecure than southern regions, while in China, coastal regions and administrative areas that contain large cities such as Beijing and Shanghai are more food secure.  Finally, in many African countries, arid regions have higher rates of moderate and severe food insecurity, such as the Sahel in Mali and Niger, the Central and Western Zones of Tanzania, and the Somali Region of Ethiopia.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a administrative area level using the FIES, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of economic growth in developing countries.  Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of moderately food insecure people.

>>>>

%\maketitle
\begin{abstract}
This paper presents the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.  Using microdata from 75 countries and filling gaps using machine learning, we find significant heterogeneity in levels of food insecurity around the world.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of severe food insecurity are declining, resulting in global decreases in the total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, we predict it will continue to increase through the end of the 2020s.  Overall, we conclude that current trends in development and demographic change will still leave a large share of the world's population still experiencing hunger by 2030.  \end{abstract}

@@

This paper contributes to shifting the research frontier by filling these two gaps in terms of spatial coverage and forecasts by developing a machine learning model of FIES-based food insecurity rates based on covariates that are available for territories where actual survey data are not available. This model is then employed to obtain forecasts to 2030 and assess likely future trends in food insecurity. We take advantage of individual-level characteristics collected in FAO's FIES surveys to push the analysis beyond the country level to a sub-national level. To our knowledge, this is the first effort to model FIES-based food insecurity at that scale.

@@

\section{Methods}
\subsection{Disaggregation}
The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  We use a similar methodology and calculate post-stratification weights at a subnational level.  We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.  This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.

\subsection{Covariates}
We model subnational rates of moderate and severe food insecurity as a function of several covariates related to human development, nutrition, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).  We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.  For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.

Many of the variables require the harmonization of subnational historical data with projections at the national level.  These include mean years of schooling, GDP per capita, as well as population. We use observed trajectories in the historical distribution of variables among subnational areas within a country to predict the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.

For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions for variables up to 2030. We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.  This method is used to obtain forecasts of the rates of stunting and wasting up to the year 2030.

@@

\subsection{Modeling}
We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.  A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.  By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}. Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.  We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.

@@

When performing cross validation, it is typical to select observations at random.  However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.  Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.  Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.  By this validation metric, we find that our models perform well, with an $R^2$ of 0.792 and mean average error (MAE) of 0.071 for moderate food insecurity and an $R^2$ of 0.746 and a MAE of 0.041 for severe food insecurity.  After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.945 and a MAE of 0.018 for severe food insecurity.

@@

\subsection{Food Insecurity Across Space}
We find substantial heterogeneity in the global distribution of severe food insecurity across countries and subnational regions (See Fig. \ref{fig:map}).  For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with 20\% of people over the corresponding threshold in at least one subnational area in every country except Gabon, Djibouti and Equatorial Guinea.  Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.  In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10\% and 15\% in northern Brazil, many central Asian and middle-eastern countries, as well as India and Indonesia.

@@

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of economic growth in developing countries.  Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of moderately food insecure people.
